---
title: "Final Project"
author: "Jia Yu Lin"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

Importing Data

```{r}
pacman::p_load(tidyverse, magritter, data.table, R.utils)
housing = read.csv("C:\\Users\\Cetus\\Desktop\\housing_data_2016_2017.csv")

housing = as_tibble(housing)
housing

summary(housing)
#skim(housing)
```
Selecting columns
```{r}
housing_data = housing %>%
  select(approx_year_built, cats_allowed, common_charges, community_district_num, coop_condo, dining_room_type, dogs_allowed, fuel_type, full_address_or_zip_code, garage_exists, kitchen_type, maintenance_cost, num_bedrooms, num_floors_in_building, num_full_bathrooms, num_half_bathrooms, num_total_rooms, parking_charges, pct_tax_deductibl, sq_footage, total_taxes, walk_score, sale_price)

housing_data

summary(housing_data)
```

Checking for NAs

```{r}
housing_data %>% 
  summarise_all(funs(sum(is.na(.))))
```
Inspecting data
```{r}
# approx_year_built
housing_data %>%
  count(approx_year_built)

#percentage of approx_year_built
# year_built_percentage = table(housing_data$approx_year_built)
# prop.table(year_built_percentage)


# cats_allowed
housing_data %>%
  count(cats_allowed)

#percentage of cats_allowed
# cats_allowed_percentage = table(housing_data$cats_allowed)
# prop.table(cats_allowed_percentage)

# common_charges
housing_data %>%
  count(common_charges)

#community_district_num
housing_data %>%
  count(community_district_num)

# community_district_percentage = table(housing_data$community_district_num)
# prop.table(community_district_percentage)

# coop_condo
housing_data %>%
  count(coop_condo)

# dining_room_type
housing_data %>%
  count(dining_room_type)

# dogs_allowed
housing_data %>%
  count(dogs_allowed)

# fuel_type
housing_data %>%
  count(fuel_type)

# full_address_or_zip_code
housing_data %>%
  count(full_address_or_zip_code)

# garage_exist
housing_data %>%
  count(garage_exists)

# kitchen_type
housing_data %>%
  count(kitchen_type)

# maintenance_cost
housing_data %>%
  count(maintenance_cost)

# num_bedrooms
housing_data %>%
  count(num_bedrooms)

# num_floors_in_building
housing_data %>%
  count(num_floors_in_building)

# num_full_bathrooms
housing_data %>%
  count(num_full_bathrooms)

# num_half_bathrooms
housing_data %>%
  count(num_half_bathrooms)

# num_total_rooms
housing_data %>%
  count(num_total_rooms)

# parking_charges
housing_data %>%
  count(parking_charges)

# pct_tax_deductibl
housing_data %>%
  count(pct_tax_deductibl)

# sq_footage
housing_data %>%
  count(sq_footage)

# total_taxes
housing_data %>%
  count(total_taxes)

# walk_score
housing_data %>%
  count(walk_score)

# sale_price
housing_data %>%
  count(sale_price)
```
```{r}
# kitchen_type_percentage = table(housing_data$kitchen_type , exclude=NULL)
# prop.table(kitchen_type_percentage)
# housing_data %>% group_by(sq_footage)%>%
#   summarise(mean= mean(sq_footage), sd= sd(sq_footage), max = max(sq_footage),min = min(sq_footage))
c(mean = mean(housing_data$maintenance_cost , na.rm = TRUE ), sd= sd(housing_data$maintenance_cost ,  na.rm = TRUE ), max = max(housing_data$maintenance_cost , na.rm = TRUE ),min = min(housing_data$maintenance_cost ,  na.rm = TRUE ) )
# library(skimr)
# skim(housing_data)
```

Turning `dogs_allowed`, `cats_allowed`, `garage_exists` into binary
```{r}
# dogs_allowed
housing_data = housing_data %>%
  mutate(dogs_allowed = if_else(dogs_allowed == 'no', 0, 1))

housing_data$dogs_allowed = as.factor(housing_data$dogs_allowed)

# cats_allowed
housing_data = housing_data %>%
  mutate(cats_allowed = if_else(cats_allowed == 'no', 0, 1))

housing_data$cats_allowed = as.factor(housing_data$cats_allowed)

# garage_exists
housing_data = housing_data %>%
  mutate(garage_exists = if_else(is.na(garage_exists), 0, 1))

housing_data$garage_exists = as.factor(housing_data$garage_exists)

#sanity check
housing_data
```

Turning `coop_condo`, `dining_room_type`, `fuel_type`, `kitchen_type` into categorical factors
```{r}

# coop_condo
housing_data = housing_data %>%
  mutate(coop_condo = as.factor(coop_condo))
  
# dining_room_type
housing_data = housing_data %>%
  mutate(dining_room_type = as.factor(dining_room_type))

# fuel_type
tabulate = sort(table(housing_data$fuel_type))

housing_data$fuel_type = as.character(housing_data$fuel_type)

housing_data$fuel_type = ifelse(housing_data$fuel_type %in% names(tabulate[tabulate < 10]), "other", housing_data$fuel_type)

housing_data$fuel_type = as.factor(housing_data$fuel_type)

# kitchen_type
             
housing_data$kitchen_type = as.character(housing_data$kitchen_type)

housing_data$kitchen_type = ifelse(tolower(substr(housing_data$kitchen_type, 1, 3)) == "com" , "combo", housing_data$kitchen_type)

housing_data$kitchen_type = ifelse(tolower(substr(housing_data$kitchen_type, 1, 3)) == "eat" , "eat in", housing_data$kitchen_type)

housing_data$kitchen_type = ifelse(tolower(substr(housing_data$kitchen_type, 1, 3)) == "eff" , "efficiency", housing_data$kitchen_type)

housing_data$kitchen_type = as.factor(housing_data$kitchen_type)

# drop 1955 row

housing_data = housing_data[housing_data$kitchen_type != "1955", ]

# sanity check
housing_data 

```


`parking_charges`, `pct_tax_deductibl`, `total_taxes`, `num_half_bathrooms`, `common_charges`, `maintenance_cost` turned into numeric
```{r}
# parking_charges
housing_data$parking_charges = as.numeric(gsub("\\$", "", housing_data$parking_charges))

# pct_tax_deductibl

# total_taxes
housing_data$total_taxes = as.numeric(gsub("\\$", "", housing_data$total_taxes))

# num_half_bathrooms


# common_charges
housing_data$common_charges = as.numeric(gsub("\\$", "", housing_data$common_charges))

# maintenance_cost
housing_data$maintenance_cost = as.numeric(gsub("\\$", "", housing_data$maintenance_cost))


housing_data
```

Cleaning up `sale_price`
```{r}
housing_data$sale_price = as.numeric(gsub("[\\$,]", "", housing_data$sale_price))

housing_data = housing_data[!is.na(housing_data$sale_price), ]

housing_data
```

Split X and y
```{r}
# full_address_or_zip_code taken out
X = housing_data %>% select(approx_year_built, cats_allowed, common_charges, community_district_num, coop_condo, dining_room_type, dogs_allowed, fuel_type, garage_exists, kitchen_type, maintenance_cost, num_bedrooms, num_floors_in_building, num_full_bathrooms, num_half_bathrooms, num_total_rooms, parking_charges, pct_tax_deductibl, sq_footage, total_taxes, walk_score)

y = housing_data$sale_price
```


Missingness
```{r}

pacman::p_load(missForest)

XimpMF = missForest(data.frame(X))$ximp

XimpMF
```


```{r}
XimpMF$sale_price = y
```


Loading other data
```{r}
# # This is for calculating distances with lat-long
# pacman::p_load(geosphere)
# 
# # This is for using google map API or whatever
# pacman::p_load(ggmap)
# 
# # WORKS TO PULL LAT-LONG
# register_google(key = "u need api key")
# pacman::p_load(ggmap)
# locs <- housing_data[sample(1000),]$full_address_or_zip_code
# newlocs = geocode(locs, output = "latlon")
# newlocs
# ?register_google
# 
# pacman::p_load(geosphere)
# # dist array is just me testing distances using the function
# for (i in 1:1000) {
#   dist_array[i] = distm(midtown_coords, c(abs(newlocs$lat[i]), abs(newlocs$lon[i])), fun = distHaversine) / 1609 # divide by 1609 to turn into miles
# }
```

Splitting data for training
```{r}
seed = 1984
set.seed(seed)
K = 5
test_prop = 1 / K
train_indices = sample(1 : nrow(XimpMF), round((1 - test_prop) * nrow(XimpMF)))
XimpMF_train = XimpMF[train_indices, ]
y_train = XimpMF_train$sale_price
X_train = XimpMF_train
X_train$sale_price = NULL

test_indices = setdiff(1 : nrow(XimpMF), train_indices)
XimpMF_test = XimpMF[test_indices, ]
y_test = XimpMF_test$sale_price
X_test = XimpMF_test
X_test$sale_price = NULL

```


Loading in YARF
```{r}
if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
options(java.parameters = "-Xmx4000m")
pacman::p_load(YARF)
```

Regression Tree (training)
```{r}
reg_tree_mod = YARFCART(X_train, y_train)
illustrate_trees(reg_tree_mod, max_depth = 5, length_in_px_per_half_split = 30, open_file = TRUE)

```

OLS (training)
```{r}
OLS = lm(y_train ~ ., data.frame(X_train))

summary(OLS)
summary(OLS)$r.squared

#RMSE
sd(y_train - OLS$fitted.values)

```


OLS (testing)
```{r}
y_hat_test_linear = predict(OLS, X_test)
e = y_test - y_hat_test_linear

#RMSe
sd(e)

#r^2s
1 - sd(e) / sd(y_test)
```




RF

```{r}
num_trees = 500
mod_rf_2 = YARF(X_train, y_train, num_trees = num_trees, seed = seed, calculate_oob_error = FALSE, mtry = 7)
#summary(mod_rf_2)
mod_rf_1 = YARF(X_train, y_train)
summary(mod_rf_1)

rmse_rf = sd(y_test - predict(mod_rf_1, X_test))
#rmse_rf =sqrt(mean((y_test - predict(mod_rf, X_test))^2))
rmse_rf

1 - rmse_rf / sd(y_test)

illustrate_trees(mod_rf_1, max_depth = 5, length_in_px_per_half_split = 30, open_file = TRUE)
```